---
output: html_document
---

```{r setup, include=FALSE}
# load libraries
library(tidyverse)
library(conflicted)
library(viridis)
library(magrittr)
library(cowplot)
library(ggdendro)

# configure knit settings
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 4)

# resolve package conflicts
filter <- dplyr::filter
select <- dplyr::select
```

<br>

## Hierarchical Clustering

<br>

#### Prep

In order to make, `ggplot` plots of hierarchical cluster, we'll need the `ggplot` extensions `ggdendro`. If you don't already have it installed, uncomment the code in the chunk below and install it now.

```{r}
#install.packages('ggdendro')
```

<br>

We're going to use some simulated data from a normal distribution to demo hierarchical clustering. Run the code in the chunk below before continuing.

```{r}
set.seed(42)
tibble(x = rnorm(6, mean = 5, sd = 2), 
       y = x + runif(6),
       label = c('A', 'B', 'C', 'D', 'E', 'F')) -> clust_demo
```

<br><br>

### Getting the data in the correct format with `dist()`

The clustering algorithm requires a distance matrix. A distance matrix calculates the euclidean distance between every row in the table and returns a distance matrix with the distances between all rows.

<br>

We need to convert our data into a distance matrix using the function `dist()`

```{r}
clust_demo %>% 
# make the tibble into a dataframe because we need to make the labels
# into rownames and tibbles don't allow rownames
  as.data.frame() %>%
# turn the data labels into rownames so they're carried through the distance
# matrix and hierarchical clustering calculations
  column_to_rownames('label') %>%
# use dist() to calculate a distance matrix
  dist(.) -> clust_demo_dist
```

<br>

### Calculate the clustering with `hclust()`

The `hclust()` function does the hierarchical clustering calculations.

```{r}
hclust(clust_demo_dist) -> clust_demo_hclust
```

<br>

### Base R plot the hierarchical clustering

You can use the base R `plot()` to directly plot the `hclust` object.

```{r}
plot(clust_demo_hclust)
```

<br>

### Plot hierarchical clustering in `ggplot` using `ggdendro`

It requires more wrangling to plot the `hclust` clustering with ggplot, but the `ggendro` package will do most of the wrangling for us.

```{r}
# start with the saved hclust object
clust_demo_hclust %>% 
# as.dendrogram() turns the hclust results into a special dendrogram class 
# that R uses for representing any kind of tree
  as.dendrogram() %>% 
# dendro_data() turns the dendrogram class data into numbers that are plottable
  dendro_data() -> clust_demo_ggdendro
```

<br>

#### `ggdendrogram()`

Plot using `ggdendro`'s helper function, `ggdendro()`. It plots the dendrogram as a ggplot object.

```{r}
ggdendrogram(clust_demo_ggdendro)
```

You can modify this like any other ggplot plot. For example, I don't like the y axis labels, so I'll remove them.

```{r}
ggdendrogram(clust_demo_ggdendro) +
  theme(axis.text.y = element_blank())
```

<br>

#### Plotting the dendrogram with straight `ggplot()`

However, you can't see (unless you look at the package code) what exact `ggplot` commands `ggdendrogram()` is plotting, which is super annoying when you want to make adjustments to the plot. However, as you can see when you look at the `dendro_data()` table below, all the numbers are there, so we can just plot it ourselves.

```{r}
clust_demo_ggdendro
```

<br>

Now we can plot it with `ggplot()`

```{r}
# for once, don't put any data in ggplot() !!!
# the stem and label information is in separate tables, so we want to supply
# separate data to separate geoms
ggplot() +
# the segments table contains the information for plotting branches, so supply
# that to geom_segment() to plot the branches of the tree
  geom_segment(data = clust_demo_ggdendro$segments, aes(x = x, y = y, xend = xend, yend = yend)) +
# the labels table has the labels for the ends of the branches, so supply that
# to geom_text() to label the ends of the branches
  geom_text(data = clust_demo_ggdendro$labels, aes(x = x, y = y, label = label), vjust = 1.25) +
  theme_dendro()
```


<br><br>

