---
title: "Examine, Subset, and Filter Data"
output: html_document
---

```{r setup, include=FALSE}
# load libraries
library(tidyverse)
library(viridis)
library(conflicted)

# configure knit settings
knitr::opts_chunk$set(echo = TRUE)

# resolve package conflicts
filter <- dplyr::filter
select <- dplyr::select
```

# Examine, Subset and Filter Data

## Examine/Look at Your Data

```{r}
iris
```

```{r}
head(iris)
```

```{r}
head(iris, 10)
```

```{r}
tail(iris)
```

```{r}
dim(iris)
```

```{r}
str(iris)
```

## Quick Detour to Logicals

These operators are how you compare values to each other. Keep them in mind for subsetting and filtering going forward.

- equals `==`
- not equal to `!=`
- greater than `>`
- greater than or equal to `>=`
- less than `<`
- less than or equal to `<=`
- and `&`
- or `|`

## Subsetting

### Basics

```{r}
iris[]
```

```{r}
# return row 3, column 5
iris[3,5]
```

```{r}
# return row 1 in all columns (no number after the comma selects all columns)
iris[1,]

# return all rows in column 1 (no number before the comma selects all rows)
iris[,1]
```

```{r}
iris$Petal.Width
```

### Other stuff you can put in square brackets

```{r}
# drop the first column
iris[,-1]
```

```{r}
# select rows 1 to 5 and columns 2 to 4
iris[1:5,2:4]
```

```{r}
iris[c(37, 49, 132, 12, 77), c(1, 3)]
```

```{r}
# get all columns and all rows where sepal length is greater than or equal to 5
iris[iris$Sepal.Length >= 5,]

# get all columns and all rows species is NOT viriginica
iris[iris$Species != 'virginica',]
```

## Filtering And Selecting

```{r}
# load just the dplyr package
library(dplyr)

### OR

# load the tidyverse
library(tidyverse)
```

### `filter()`

```{r}
filter(iris, Petal.Length <= 2)
```

```{r}
filter(iris, Petal.Length >= 2, Sepal.Length <= 5)
```

### `select()`

```{r}
select(iris, Species)
```

```{r}
select(iris, Sepal.Width, Petal.Width, Species)
```

```{r}
select(iris, -Species)
```


