---
title: 'Examine, Subset, and Filter Data: Week 2 Practice'
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
# load libraries
library(tidyverse)
library(viridis)
library(conflicted)

# configure knit settings
knitr::opts_chunk$set(echo = TRUE)

# resolve package conflicts
filter <- dplyr::filter
select <- dplyr::select
```

# Examine, Subset and Filter Data
### Week 2 Practice

We'll use the built-in dataset ToothGrowth for this practice.

## Examine Data

Look at the `head()` of ToothGrowth

```{r}
head(ToothGrowth)
```

And the `tail()`

```{r}
tail(ToothGrowth)
```

What are the dimensions of ToothGrowth?

```{r}
dim(ToothGrowth)
```

What is ToothGrowth's structure?

```{r}
str(ToothGrowth)
```

## Subset Data

Select the first column only using square brackets.

```{r}
ToothGrowth[,1]
```

Select the 42nd row using square brackets.

```{r}
ToothGrowth[42,]
```

Select rows 20 to 40 and columns 1 and 3 using square brackets.

```{r}
ToothGrowth[20:40, c(1,3)]
```

Select the supp column using a dollar sign

```{r}
ToothGrowth$supp
```

## Filter/Select

Filter ToothGrowth for a dose greater than 1

```{r}
filter(ToothGrowth, dose > 1)
```

Filter ToothGrowth for a supplements that are NOT "VC" and length less than or equal to ten

```{r}
filter(ToothGrowth, supp != 'VC', len < 10)
```

Select all the columns, BUT the dose columns

```{r}
select(ToothGrowth, len, supp)

### OR

select(ToothGrowth, -dose)
```


## Anything Goes Challenge Questions

Return only VC supplements, with length less than or equal to 10 and greater than or equal to 25.

```{r}
ToothGrowth[(ToothGrowth$supp == 'VC' & (ToothGrowth$len <= 10 | ToothGrowth$len >= 25)),]

### OR

filter(ToothGrowth, supp == 'VC', len <= 10 | len >= 25)
```

Return length less than 10 or dose equals 2 and drop the supp column

```{r}
ToothGrowth[ToothGrowth$len < 10 | ToothGrowth$dose == 2, c(1, 3)]

### OR

select(filter(ToothGrowth, len < 10 | dose == 2), -supp)

### OR

filter(ToothGrowth, len < 10 | dose == 2) %>% select(-supp)
```





