---
title: 'Examine, Subset, and Filter Data: Week 2 Practice'
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
# load libraries
library(tidyverse)
library(viridis)
library(conflicted)

# configure knit settings
knitr::opts_chunk$set(echo = TRUE)

# resolve package conflicts
filter <- dplyr::filter
select <- dplyr::select
```

# Examine, Subset and Filter Data
### Week 2 Practice

We'll use the built-in dataset ToothGrowth for this practice.

## Examine Data

Look at the `head()` of ToothGrowth

```{r}

```

And the `tail()`

```{r}

```

What are the dimensions of ToothGrowth?

```{r}

```

What is ToothGrowth's structure?

```{r}

```

## Subset Data

Select the first column only using square brackets.

```{r}

```

Select the 42nd row using square brackets.

```{r}

```

Select rows 20 to 40 and columns 1 and 3 using square brackets.

```{r}

```

Select the supp column using a dollar sign

```{r}

```

## Filter/Select

Filter ToothGrowth for a dose greater than 1

```{r}

```

Filter ToothGrowth for a supplements that are NOT "VC" and length less than or equal to ten

```{r}

```

Select all the columns, BUT the dose columns

```{r}

```


## Anything Goes Challenge Questions

Return only VC supplements, with length less than or equal to 10 and greater than or equal to 25.

```{r}

```

Return length less than 10 or dose equals 2 and drop the supp column

```{r}

```





