---
title: "Practice: dplyr() Data Manipulation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(viridis)
library(conflicted)
library(nycflights13)

filter <- dplyr::filter
```

# Practice: `dplyr()` Data Manipulation

We'll use another data table from the nycflights13 library, airports. Don't forget to look at the table before doing anything!

```{r}
airports %>% head()
```

## `dplyr()` functions

### `filter()`

Filter airports for locations in our time zone, the US Eastern time zone.

```{r}
airports %>% filter(tz == -5)

### OR

airports %>% filter(tzone == 'America/New_York')
```

### `select()`

Select all columns BUT latitude and longitude.

```{r}
airports %>% select(-lat, -lon)

### OR

airports %>% select(faa, name, alt, tz, dst, tzone)

### OR

airports %>% select(-(lat:lon))

### OR

airports %>% select(faa, name, alt:tzone)
```

### `mutate()`

Add a column indicating the year the data was collected, 2013

```{r}
airports %>% mutate(dat_collect = 2013)
```

### `arrange()`

Arrange the table by latitude, then longitude

```{r}
airports %>% arrange(lat, lon)
```

### `group_by()` / `summarize()`

Find the average altitude by time zone.

```{r}
airports %>% group_by(tzone) %>% summarize(avg_alt = mean(alt))
```

## Chain functions together with pipes

Filter for airports with altitude above 2000 feet and select their FAA airport codes and names

```{r}
airports %>% filter(alt > 2000) %>% select(faa, name)
```

Filter for airports with an altitude less than 1000 feet and find their average latitude and longitudes by timezone.

```{r}
airports %>% filter(alt < 1000) %>% group_by(tzone) %>% summarize(avg_lat = mean(lat), avg_lon = mean(lon))
```

Filter for latitude between 35 and 50, then plot boxplots of altitude by time zone.

```{r}
airports %>% filter((lat >= 35 & lat <= 50)) %>% ggplot(aes(x = tzone, y = alt)) + geom_boxplot()
```




