---
output: html_document
---

```{r setup, include=FALSE}
# load libraries
library(tidyverse)
library(conflicted)
library(viridis)
library(broom)
library(magrittr)

# configure knit settings
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 4)

# resolve package conflicts
filter <- dplyr::filter
select <- dplyr::select
```

# Week 8 practice: Linear modeling and `broom`

To practice the new tests we've learned, we'll use the built-in `mtcars` dataset. 

## Test Practice

### correlation

What is the correlation between horsepower and weight? Is it significant?

```{r}
### answer here
```

<br>

### linear model

#### Plot

Plot a scatterplot of the rear axle ratio vs. fuel displacement. Does the relationship look linear?

```{r}
### answer here 
```

<br>

#### Test

Since the relationship looks approximately linear, test it with `lm()`

```{r}
### answer here
```

<br>

### ANOVA

Do an ANOVA of quarter mile time over the number of cylinders in the the engine.

```{r}
### answer here)
```

<br>

Now take the same ANOVA from the chunk above and add a post-hoc Tukey test to it.

```{r}
### answer here
```

<br>

### broom

Pick one of the three tests you did above and practice tidying it up with the three `broom` functions!

<br>

#### `tidy()`

Add `tidy()` onto end of your test.

```{r}
### answer here
```

<br>

#### `glance()`

Add `glance()` onto the end of your test to look at the model parameters.

```{r}
### answer here
```

<br>

#### `augment()`

Add the results of the test back to mtcars using `augment()`

```{r}
### answer heres)
```

<br>

Use the augmented table from the chunk above to plot something.

```{r}
### answer here
```

<br><br>

## Come Up with Your Own Questions

Use the tests talked about this week, `cor()`, `cor.test()`, `lm()`, `aov()`, and/or `TukeyHSD()` to ask two more questions about the `mtcars` dataset.

<br>

### Question 1

#### Plot

Look at mtcars again. What's another question you could ask about it? Plot the variable(s) you're interested in below

```{r}
### answer here
```

<br>

#### Ask a Question

Based on the variables you just plotted, what's your question?

**Type your question here** 

<br>

#### Test

Pick one of the tests to answer your question and run it in the chunk below.

```{r}
### answer here
```

<br>

#### Answer the Question

What do you conclude from your test?

**Type your answer here** 

<br>

### Question 2

### Plot

Look at mtcars again. What's another question you could ask about it? Plot the variable(s) you're interested in below

```{r}
### answer here
```

<br>

#### Ask a Question

Based on the variables you just plotted, what's your question?

**Type your question here** Can displacement predict horsepower?

<br>

#### Test

Pick one of the tests to answer your question and run it in the chunk below.

```{r}
### answer here
```

<br>

#### Answer the Question

What do you conclude from your test?

**Type your answer here** 


<br><br>







