---
output: html_document
---

```{r setup, include=FALSE}
# load libraries
library(tidyverse)
library(conflicted)
library(viridis)
library(broom)
library(cowplot)

# configure knit settings
knitr::opts_chunk$set(echo = TRUE)

# resolve package conflicts
filter <- dplyr::filter
select <- dplyr::select
ggsave <- cowplot::ggsave
```

## Correlation Review

Correlation gives you the direction and strength of the linear association between two numeric variables. Usually referred to using the variable r, correlation values lie between -1 and 1.

#### Example 1: Perfect Correlation

```{r}
# get random numbers to plot
runif(100, -10, 10) %>% round() %>% as_tibble() -> random
```

```{r}
# plot perfect correlation
ggplot(random, aes(x = value, y = value)) +
  geom_point() + 
  theme(axis.title = element_blank()) -> perfect_cor

# perfect inverse correlation
ggplot(random, aes(x = value, y = -value)) +
  geom_point() + 
  theme(axis.title = element_blank()) -> perfect_inverse_cor

# plot together
plot_grid(perfect_cor, perfect_inverse_cor)

# save
#ggsave('perfect_cor.png', width = 4, height = 2)
```

#### Example 2: Some plots with different correlations

Generate random numbers to make correlations

```{r}
tibble(random1 = rnorm(100, 1, 1),
       random2 = rnorm(100, 2, 0.5)) -> random_norm
```

Create correlations using the random numbers 

```{R}
random_norm %>% mutate(cor_high = (0.8*random1 + sqrt(1-(0.8^2))*random2),
                       cor_no = (-0.05*random1 + sqrt(1-(-0.05^2))*random2),
                       cor_mod = (-0.4*random1 + sqrt(1-(-0.4^2))*random2),
                       cor_low = (0.1*random1 + sqrt(1-(0.1^2))*random2)) -> random_cor
```

Get correlation values

```{r}
cor(random_cor)
```

Plot the correlations

```{r}
# high correlation
ggplot(random_cor, aes(x = random1, y = cor_high)) + 
  geom_point() +
  theme(axis.title = element_blank()) -> cor_high

# moderate correlation
ggplot(random_cor, aes(x = random1, y = cor_mod)) + 
  geom_point() +
  theme(axis.title = element_blank()) -> cor_mod

# low correlation
ggplot(random_cor, aes(x = random1, y = cor_low)) + 
  geom_point() +
  theme(axis.title = element_blank()) -> cor_low

# no correlation
ggplot(random_cor, aes(x = random1, y = random2)) + 
  geom_point() +
  theme(axis.title = element_blank()) -> cor_no

# plot together
plot_grid(cor_high, cor_mod, cor_low, cor_no)

# save
Eggsave('different_cors.png', width = 4, height = 4)
```

#### Example 3: Effect of Variability on Correlation

Get different variabilties

```{r}
random_norm %>% mutate(var1 = (0.9*random1 + sqrt(1-(0.8^2))*random2),
                       var2 = (0.7*random1 + sqrt(1-(0.7^2))*random2),
                       var3 = (0.5*random1 + sqrt(1-(0.6^2))*random2),
                       var4 = (0.3*random1 + sqrt(1-(0.5^2))*random2)) -> random_var
```

Find the correlation of the varied numbers

```{r}
cor(random_var)
```

Plot the differences in variation

```{r}
# high correlation
ggplot(random_var, aes(x = random1, y = var1)) + 
  geom_point() +
  theme(axis.title = element_blank()) -> var1

# moderate correlation
ggplot(random_var, aes(x = random1, y = var2)) + 
  geom_point() +
  theme(axis.title = element_blank()) -> var2

# low correlation
ggplot(random_var, aes(x = random1, y = var3)) + 
  geom_point() +
  theme(axis.title = element_blank()) -> var3

# no correlation
ggplot(random_var, aes(x = random1, y = var4)) + 
  geom_point() +
  theme(axis.title = element_blank()) -> var4

# plot together
plot_grid(var1, var2, var3, var4)

#ggsave('var_on_cor.png', width = 4, height = 4)
```

## Linear Regression Review

### Example 4: Linear Regression and Correlation

```{r}
random_var %>% select(random1, var4) %>% .[1:25,] -> lm_examp
```

```{r}
lm(var4 ~ random1, data = lm_examp) %>% tidy()
```

```{r}
# no line of fit
ggplot(lm_examp, aes(x = random1, y = var4)) +
  geom_point(size = 3, color = 'red2') +
  theme_classic() + 
  theme(axis.title = element_blank()) -> no_fit

# very bad line of fit
ggplot(lm_examp, aes(x = random1, y = var4)) +
  geom_abline(slope = 0.2, intercept = 1.5) +
  geom_segment(aes(x = random1, xend = random1, y = var4, yend = 0.2*random1 + 1.5)) +
  geom_point(size = 3, color = 'red2') +
  theme_classic() + 
  theme(axis.title = element_blank()) -> bad_fit

# better line of fit
ggplot(lm_examp, aes(x = random1, y = var4)) +
  geom_abline(slope = 0.5, intercept = 1.5) +
  geom_segment(aes(x = random1, xend = random1, y = var4, yend = 0.5*random1 + 1.5)) +
  geom_point(size = 3, color = 'red2') +
  theme_classic() + 
  theme(axis.title = element_blank()) -> better_fit

# line of best fit
ggplot(lm_examp, aes(x = random1, y = var4)) +
  geom_abline(slope = 0.323, intercept = 1.86) +
  geom_segment(aes(x = random1, xend = random1, y = var4, yend = 0.323*random1 + 1.86)) +
  geom_point(size = 3, color = 'red2') +
  theme_classic() + 
  theme(axis.title = element_blank()) -> best_fit

# plot together
plot_grid(no_fit, bad_fit, better_fit, best_fit)

# save
#ggsave('linear_regression.png', width = 4, height = 4)
```







