```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 4)

library(tidyverse)
library(viridis)
library(conflicted)
library(nycflights13)

filter <- dplyr::filter
```

# Practice: `dplyr()` Data Manipulation

We'll use another data table from the nycflights13 library, airports. Don't forget to look at the table before doing anything!

```{r}
airports
```

## `dplyr()` functions

### `filter()`

Filter airports for locations in our time zone, the US Eastern time zone.

```{r}
filter(airports, tz == -5)
```

### `select()`

Select all columns BUT latitude and longitude.

```{r}
select(airports, -lat, -lon)
```

### `mutate()`

Add a column indicating the year the data was collected, 2013

```{r}
# wrong, use only one equal sign when creating a new column because we're *assigning* (=) new values, not checking if values are equal (==)
airports %>% mutate(collected == 2013)

airports %>% mutate(collected = 2013)
```

### `arrange()`

Arrange the table by latitude, then longitude

```{r}
arrange(airports, lat, lon)

### OR equivalent answer

airports %>% arrange(lat, lon)
```

### `group_by()` / `summarize()`

Find the average altitude by time zone.

```{r}
airports %>% group_by(tzone) %>% summarize(average_alt = mean(alt))
```

## Chain functions together with pipes

Filter for airports with altitude above 2000 feet and select their FAA airport codes and names

```{r}
airports %>% filter(alt >= 2000) %>% select(faa, name)
```

Filter for airports with an altitude less than 1000 feet and find their average latitude and longitudes by timezone.

```{r}
airports %>% filter(alt < 1000) %>% group_by(tzone) %>% summarize(avg_lat = mean(lat), avg_lon = mean(lon))
```

Filter for latitude between 35 and 50, then plot boxplots of altitude by time zone.

```{r}
# class answer
airports %>% filter(lat > 35 & lat < 50) %>%
  ggplot() + geom_boxplot(aes(y = alt, x = tzone))

# doesn't work -- best practice is to put aesthetics (aes()) inside ggplot() so all geoms can inherit the aesthetics (see the data) from ggplot()
airports %>% filter(lat > 35 & lat < 50) %>%
  ggplot() + geom_point(aes(x = lat, y = lon)) + geom_line()

# this works!
airports %>% filter(lat > 35 & lat < 50) %>%
  ggplot(aes(x = lat, y = lon)) + geom_point() + geom_line()


# example of how color works differently when you put in ggplot(aes()) vs. geom_*(aes())
airports %>% filter(lat > 35 & lat < 50) %>%
  ggplot(aes(x = lat, y = lon)) + geom_point(aes(color=tzone)) + geom_line()

airports %>% filter(lat > 35 & lat < 50) %>%
  ggplot(aes(x = lat, y = lon, color = tzone)) + geom_point() + geom_line()
```




